---
title: "MT5763_1_200030711"
author: "Alexander Curtis"
date: "03/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:
```{r, message = FALSE}
library(tidyverse)
library(lubridate)
library(gridExtra)
```

## Data wrangling

Data wrangling for the Seoul dataset:
```{r}
BikeSeoul <- read.csv("BikeSeoul.csv") %>% 
  # Remove unwanted columns
  select(-Visibility..10m.,
         -Dew.point.temperature.C.,
         -Solar.Radiation..MJ.m2., -Rainfall.mm.,
         -Snowfall..cm.) %>%
  # Keep only functioning days then remove column
  filter(Functioning.Day == "Yes") %>%
  select(-Functioning.Day) %>%
  rename(Count = Rented.Bike.Count, 
         Temperature = Temperature.C.,
         Humidity = Humidity...,
         WindSpeed = Wind.speed..m.s.,
         Season = Seasons) %>%
  mutate(Date = as.Date(Date, "%d/%m/%Y")) %>%
  mutate(FullDate = make_datetime(year = year(Date),
                                  month = month(Date),
                                  day = day(Date),
                                  hour = Hour)) %>%
  mutate(Holiday = factor(Holiday, 
                          levels = c("Holiday", "No Holiday"),
                          labels = c("Yes", "No"))) %>%
  mutate(Season = factor(Season, 
                         levels = c("Spring", "Summer", "Autumn", "Winter")))
```


Data wrangling for the Washington DC dataset:
```{r}
BikeWashingtonDC <- read.csv("BikeWashingtonDC.csv") %>%
  # Remove unwanted columns
  select(-instant,
         -yr,
         -mnth,
         -weekday,
         -workingday,
         -weathersit,
         -atemp,
         -casual,
         -registered) %>%
  rename(Date = dteday,
         Season = season,
         Hour = hr,
         Holiday = holiday,
         Temperature = temp,
         Humidity = hum,
         WindSpeed = windspeed,
         Count = cnt) %>%
  # Transform humidity into a percentage
  mutate(Humidity = Humidity * 100) %>%
  # Transform Temperature into degrees Celcius
  mutate(Temperature = 47 * Temperature - 8) %>%
  # Transform Windspeed into m/s
  mutate(WindSpeed = WindSpeed * 67) %>%
  mutate(Season = factor(Season,
                         levels = c(2, 3, 4, 1),
                         labels = c("Spring", "Summer", "Autumn", "Winter"))) %>%
  mutate(Holiday = factor(Holiday,
                          levels = c(1, 0),
                          labels = c("Yes", "No"))) %>%
  mutate(Date = as.Date(Date, "%Y-%m-%d")) %>%
  mutate(FullDate = make_datetime(year = year(Date),
                                  month = month(Date),
                                  day = day(Date),
                                  hour = Hour))
```


## Data Visualisation

### How does air temperature vary over the course of a year?

To get a good sense of how the air temperature changes, a plot has been created which shows the mean temperature for every day within the Seoul dataset.
```{r, message = FALSE}
BikeSeoul %>% 
  group_by(Date) %>%
  summarise(DailyMean = mean(Temperature)) %>%
  ggplot() +
  geom_line(aes(x = Date, y = DailyMean)) +
  ylab("Temperature (ºC)") +
  ggtitle("Mean Daily Air Temperature in Seoul")
```
This plot shows that the temperature is fairly low in the winter months but then climbs steadily through the year to a peak just over 30ºC around August. The temperature then falls at a similar rate to how it climbed as it falls back to around 0ºC.
The same plot has been created for the Washington DC dataset, but with just one of the two years selected so that it is easier to compare with Seoul.
```{r, message = FALSE}
BikeWashingtonDC %>% 
  filter(year(Date) == 2011) %>%
  group_by(Date) %>%
  summarise(DailyMean = mean(Temperature)) %>%
  ggplot() +
  geom_line(aes(x = Date, y = DailyMean)) +
  ylab("Temperature (ºC)") +
  ggtitle("Mean Daily Air Temperature in Washington DC")
```
The Washington plot follows a very similar trend to the one for Seoul. It also climbs steadily from 0ºC around January to just over 30ºC in August and then falls back steadily. The main difference between the plots is that the daily temperature variation is a lot higher in Washington. This can be seen by the sharp peaks and troughs every few days, whereas the Seoul temperature seems a lot more consistent, especially in the later half of the year.

### Do seasons affect the average number of rented bikes?

This question is best answered by creating a bar chart of the counts for each season and comparing.
```{r, message = FALSE}
p1 <- BikeSeoul %>%
  group_by(Season) %>%
  summarise(meanCount = mean(Count)) %>%
  ggplot() + 
  geom_col(aes(x = Season, y = meanCount)) + 
  ylab("Count") +
  ggtitle("Seoul")
p2 <- BikeWashingtonDC %>%
  group_by(Season) %>%
  summarise(meanCount = mean(Count)) %>%
  ggplot() + 
  geom_col(aes(x = Season, y = meanCount)) + 
  ylab("Count") +
  ggtitle("Washington DC")
grid.arrange(p1, p2, ncol = 2, top = "Average Hourly Bike Rental")
```
The plots show a similar pattern in that summer is the season with the highest average number of bike rentals, and winter is the season with the lowest. Winter is particularly low for the Seoul dataset, with only a third the number of the next lowest - spring. Autumn is the second most popular season in Seoul whereas spring is in Washington DC. However it is much closer between spring and autumn in Washington DC than it is in Seoul.

### Do holidays increase or decrease the demand for rented bikes?

To compare the effect of holidays, a box-plot has been created for each dataset, each split by whether it is a holiday or not.
```{r}
p1 <- ggplot(BikeSeoul) +
  geom_boxplot(aes(x = Holiday, y = Count)) +
  ggtitle("Seoul")
p2 <- ggplot(BikeWashingtonDC) +
  geom_boxplot(aes(x = Holiday, y = Count)) +
  ggtitle("Washington DC")
grid.arrange(p1, p2, ncol = 2, top = "Hourly Bike Rental")
```
In Seoul there is a slight decrease in demand for bikes when it is a holiday. The median and upper quartile are both slightly lower for holidays but there is still a lot of crossover between the two groups. The difference in Washington DC follows the same trend with there being lower demand on holidays, but the difference is even less pronounced.

### How does the time of day affect the demand for rented bikes?

To visualise the effect of the time of day on demand for bikes, a plot has been made of the average demand for bikes every hour of the day. The plots are arranged on top of each other to compare between the two.
```{r}
p1 <- BikeSeoul %>%
  group_by(Hour) %>%
  summarise(MeanCount = mean(Count)) %>%
  ggplot() +
  geom_line(aes(x = Hour, y = MeanCount)) +
  xlab("Time of Day") +
  ylab("Demand for Bikes") +
  ggtitle("Seoul")
p2 <- BikeWashingtonDC %>%
  group_by(Hour) %>%
  summarise(MeanCount = mean(Count)) %>%
  ggplot() +
  geom_line(aes(x = Hour, y = MeanCount)) +
  xlab("Time of Day") +
  ylab("Demand for Bikes") +
  ggtitle("Washington DC")
grid.arrange(p1, p2, nrow = 2, top = "Average Hourly Demand for Bikes")
```
Both cities here display very similar trends. The lowest demand for bikes is in the early hours of the morning around 04:00 which then rises to an initial peak at 08:00 (perhaps for the commute to work). The demand then falls to a fairly steady level around midday and then peaks in the midafternoon. This peak is at 18:00 in Seoul whereas it is at 17:00 in Washington DC which perhaps points to a difference in the working hours of the day between the cities. The demand then falls back down towards the evening, although it falls slightly more steeply in Washington DC than it does in Seoul.

### Is there an association between bike demand and the three meteorological variables?

To answer this , a plot has been made for each meteorological variable to compare between the cities.

#### Air Temperature

As there are so many datapoints in each dataset, the mean count has been plotted for each temperature. The temperature was recorded to a greater degree of precision in Seoul so there are more datapoints than in Washington DC.
```{r}
p1 <- BikeSeoul %>%
  group_by(Temperature) %>%
  summarise(MeanCount = mean(Count)) %>%
  ggplot() + 
  geom_point(aes(x = Temperature, y = MeanCount)) +
  xlab("Temperature (ºC)") +
  ylab("Bike Demand") +
  ggtitle("Seoul")
p2 <- BikeWashingtonDC %>%
  group_by(Temperature) %>%
  summarise(MeanCount = mean(Count)) %>%
  ggplot() + 
  geom_point(aes(x = Temperature, y = MeanCount)) +
  xlab("Temperature (ºC)") +
  ylab("Bike Demand") +
  ggtitle("Washington DC")
grid.arrange(p1, p2, top = "Average Hourly Demand for Bikes")
```
There is a fairly linear relationship between the air temperature and the demand for rental bikes in both cities up until about 30ºC. The demand clearly dips after this point in Seoul as the temperature heads towards 40ºC, but it is slightly less clear if this is the case in Washington DC due to the precision the temperatures were recorded to.

#### Wind Speed

A similar plot for wind speed has created to the one above. Once again the data in Seoul were recorded to a slightly higher degree of precision than in Washington DC.
```{r}
p1 <- BikeSeoul %>%
  group_by(WindSpeed) %>%
  summarise(MeanCount = mean(Count)) %>%
  ggplot() + 
  geom_point(aes(x = WindSpeed, y = MeanCount)) +
  xlab("Wind Speed (m/s)") +
  ylab("Bike Demand") +
  ggtitle("Seoul")
p2 <- BikeWashingtonDC %>%
  group_by(WindSpeed) %>%
  summarise(MeanCount = mean(Count)) %>%
  ggplot() + 
  geom_point(aes(x = WindSpeed, y = MeanCount)) +
  xlab("Wind Speed (m/s)") +
  ylab("Bike Demand") +
  ggtitle("Washington DC")
grid.arrange(p1, p2, top = "Average Hourly Demand for Bikes")
```
The trend for the wind speed is less clear than it was for air temperature. Demand seems to be lowest for the low windspeeds and for the higher extremes, and highest in the middle. There is a lot more variation for the high wind speeds, however this may just be because there are less data, and so the means do not reduce the effect of the outliers as much.

#### Humidity

Again, the same plot has been created, this time with humidity along the x-axis.
```{r}
p1 <- BikeSeoul %>%
  group_by(Humidity) %>%
  summarise(MeanCount = mean(Count)) %>%
  ggplot() + 
  geom_point(aes(x = Humidity, y = MeanCount)) +
  xlab("Humidity (%)") +
  ylab("Bike Demand") +
  ggtitle("Seoul")
p2 <- BikeWashingtonDC %>%
  group_by(Humidity) %>%
  summarise(MeanCount = mean(Count)) %>%
  ggplot() + 
  geom_point(aes(x = Humidity, y = MeanCount)) +
  xlab("Humidity (%)") +
  ylab("Bike Demand") +
  ggtitle("Washington DC")
grid.arrange(p1, p2, top = "Average Hourly Demand for Bikes")
```
The plot for the Washington DC dataset shows a clear linear trend that demand for bikes drops as humidity increases from 20% up to 100%. There are a few points of low demand with low humidity but there are not much data here so it is difficult to determine if this is significant or not.
The Seoul dataset follows a slightly different pattern. There is some high demand during low humidity around 10% (although there are not many data here) and then it is fairly constant from 15% to 75%. Demand then seems to drop fairly linearly between 75% up to 100%.


## Statistical Modelling

### Fitting the models
Fitting a linear model with log count as the response variable, and season and the three meteorological variables as predictors.
```{r}
SeoulLM <- lm(log(Count) ~ Season + Temperature + Humidity + WindSpeed,
              data = BikeSeoul)
summary(SeoulLM)
```
The summary of the linear model shows that almost all predictor varaibles are statistically significant. The only one which was not was the Summer factor of the Season variable which perhaps points to summer and spring having similar effects. The estimated coefficients generally agree with the previous analysis. Namely, that winter and high humidity reduces bike demand, and high temperature and wind speed generally increase demand. However, as seen by the plots, this relationship was not entirely linear so a quadratic plot might be more useful here. The R-squared is fairly low at below 0.5, so the model is not a great fit and so predictions made using it will not be very accurate neccessarily.
The same linear model is fitted for the Washington DC data.
```{r}
WashingtonDCLM <- lm(log(Count) ~ Season + Temperature + Humidity + WindSpeed,
              data = BikeWashingtonDC)
summary(WashingtonDCLM)
```
The results of this linear model follow the same general pattern as those for the Seoul data. However, in Washington DC, all predictor variables are significant. Interestingly, the effect of summer compared to spring is negative and the effect of autumn and winter is positive, contrary to what the bar charts showed earlier. This perhaps suggests there are some confounding factors affecting this. The rest of the coefficients follow the same pattern as those for Seoul. The R-squared is much lower for this model which suggests it is an even worse fit than the Seoul model and so predictions made using this may not be very accurate.

### Confidence interval for the coefficients

A 97% confidence interval has been calculated for the regression coefficients. For the Seoul model:
```{r}
confint(SeoulLM, level = 0.97)
```
And for the Washington DC model:
```{r}
confint(WashingtonDCLM, level = 0.97)
```
These confidence intervals are calculated under the assumption that the residuals of the model are normally distributed and homoscedastic. These assumptions can be checked using the following plots.
For the Seoul model:
```{r}
plot(SeoulLM, 1)
plot(SeoulLM, 2)
```
The Q-Q plot shows that the residuals are somewhat normal (although not at the lower end), however they do appear to have fairly constant variance. Therefore, the assumptions have not violated the assumptions and the confidence intervals are fairly reliable.
Repeating for the Washington DC dataset:
```{r}
plot(WashingtonDCLM, 1)
plot(WashingtonDCLM, 2)
```
The plots here show slightly bigger violations to the assumptions, but once again they are not too bad. The confidence intervals are therefore somewhat reliable for this linear model, although perhaps less so than they are for the Seoul model.

### Making predictions

Using the linear models above, the following predictions can be made.
For the Seoul dataset:
```{r}
dfPred <- data.frame(Season = "Winter",
             Temperature = 0,
             WindSpeed = 0.5,
             Humidity = 0.2)
logPred <- predict(SeoulLM,
        newdata = dfPred,
        interval = "prediction",
        level = 0.9)
exp(logPred)
```
This is a very wide interval so would not be much use in practice. This is mainly due to the poor fit of the model, but it also wider because a prediction interval has been used rather than a confidence interval.
Doing the same for the Washington DC dataset:
```{r}
logPred <- predict(WashingtonDCLM,
        newdata = dfPred,
        interval = "prediction",
        level = 0.9)
exp(logPred)
```
This interval is also very wide for the same reasons as before. Again, it would therefore not be very useful to make decisions based on this. 
Comparing between the two models, the predictions for bike demand in Washington DC is a lot lower than it is in Seoul, however there is still quite a lot of crossover between the two intervals as they are so wide.
